ODGI: understanding pangenome graphs

### 2021-10-22 manuscript/fig_odgi_viz.tex
drafted odgi viz part

\begin{figure*}[h!]
    \begin{subfigure}{\linewidth}
        \caption{}
        \centering
        % include first image
        \includegraphics[width=1.0\linewidth, trim=0 +2cm 0 0]{fig/visualization_1D/chr6.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.grch38_chr6_31889244-32095493.0.sorted}
        \label{fig:odgi_viz_default}
    \end{subfigure}
    \begin{subfigure}{\linewidth}
        \caption{}
        \centering
        % include second image
        \includegraphics[width=1.0\linewidth, trim=0 +2cm 0 0]{fig/visualization_1D/chr6.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.grch38_chr6_31889244-32095493.0.sorted.du}
        \label{fig:odgi_viz_color_by_path_pos}
    \end{subfigure}
    \begin{subfigure}{\linewidth}
        \caption{}
        \centering
        % include second image
        \includegraphics[width=\linewidth, trim=0 +2cm 0 0]{fig/visualization_1D/chr6.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.grch38_chr6_31889244-32095493.0.sorted.z}
        \label{fig:odgi_viz_color_by_inversion_rate}
    \end{subfigure}
    \begin{subfigure}{1\linewidth}
        \caption{}
        \centering
        % include fourth image
        \includegraphics[width=\linewidth, trim=0 0 0 0]{fig/visualization_1D/chr6.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.grch38_chr6_31889244-32095493.0.sorted.m}
        \label{fig:odgi_viz_color_by_path_depth}
    \end{subfigure}
    \caption{Visualizing the complement component 4 (C4) \textit{locus} from a human pangenome graph of 90 haplotypes: \textbf{(a)} standard mode: explain the bio stuff. \textbf{(b)} odgi viz's color by path position: explain the bio stuff XXX. \textbf{(c)} odgi viz's color by strandness: explain the bio stuff XXX. `textbf{(d)}  odgi viz's color by coverage: explain the bio stuff XXX.}
    \label{fig:odgi_viz}
\end{figure*}

>>>>

\begin{figure*}[h!]
    \begin{subfigure}{\linewidth}
        \caption{}
        \centering
        % include first image
        \includegraphics[width=1.0\linewidth, trim=0 +2cm 0 0]{fig/visualization_1D/chr6.pan.fa.c3d3224.7748b33.395c7f4.smooth.gfa.C4.sorted}
        \label{fig:odgi_viz_default}
    \end{subfigure}
    \begin{subfigure}{\linewidth}
        \caption{}
        \centering
        % include second image
        \includegraphics[width=1.0\linewidth, trim=0 +2cm 0 0]{fig/visualization_1D/chr6.pan.fa.c3d3224.7748b33.395c7f4.smooth.gfa.C4.sorted.du}
        \label{fig:odgi_viz_color_by_path_pos}
    \end{subfigure}
    \begin{subfigure}{\linewidth}
        \caption{}
        \centering
        % include second image
        \includegraphics[width=\linewidth, trim=0 +2cm 0 0]{fig/visualization_1D/chr6.pan.fa.c3d3224.7748b33.395c7f4.smooth.gfa.C4.sorted.z}
        \label{fig:odgi_viz_color_by_inversion_rate}
    \end{subfigure}
    \begin{subfigure}{1\linewidth}
        \caption{}
        \centering
        % include fourth image
        \includegraphics[width=\linewidth, trim=0 1.5cm 0 0]{fig/visualization_1D/chr6.pan.fa.c3d3224.7748b33.395c7f4.smooth.gfa.C4.sorted.m}
        \label{fig:odgi_viz_color_by_path_depth}
    \end{subfigure}
    \caption{
        Visualizing the complement component 4 (C4) pangenome graph extracted from a whole human pangenome graph of 90 haplotypes.
        In all visualizations, 10 paths are displayed: 2 reference genomes (chm13 and grch38 on the top) and 8 haplotypes from 4 individuals.
        \textbf{(a)} default modality: the image shows a quite linear graph.
        The 2 longer links at the bottom indicate the presence of a structural variant (long link) with another structural variant nested inside it (short link).
        Indeed, human C4 exists as 2 functionally distinct genes, C4A and C4B, which both vary in structure and copy number \citep{Sekar_2016}. The longer link indicates that the copy number status varies across the haplotypes represented in the pangenome.
        Moreover, C4A and C4B genes segregate in both long and short genomic forms, distinguished by the presence or absence of a human endogenous retroviral (HERV) sequence, as also highlighted by the short nested link on the bottom.
        \textbf{(b)} color by path position: the color gradients are smooths, highlighting that its node are well sorted in 1 dimension.
        The top two reference genomes and 2 haplotypes (with names starting with HG01978\#1 and HG03492\#2) go from left to right, while 5 haplotypes go in the opposite direction, as indicated by the black color on their left.
        \textbf{(c)} color by strandness: the red paths indicate the haplotypes that were assembled in reverse with respect to the 2 reference genomes.
        \textbf{(d)} color by path depth: using the Spectra color palette with 4 level of path depths, white indicates no depth, while grey, red, and yellow indicate depth 1, 2, and 3, respectively.
        Coloring by path depth, we can see that the two references present two different allele copies of the C4 genes, both of them including the HERV sequence. The entirely grey paths have one copy of these genes.
        The path with name starting with HG01071\#2 presents 3 copies of the genes (indicated by the orange color), of which only 1 with the HERV sequence (gray color in the middle of the orange region).
    }
    \label{fig:odgi_viz}
\end{figure*}

### 2021-10-22 manuscript/main.tex
compress images putting subcaptions on the left
drafted odgi viz part

\usepackage[draft]{hyperref}
\usepackage{makecell}
\usepackage{comment}
\usepackage{subcaption}
\usepackage{algorithm2e}
\usepackage[usenames,dvipsnames]{xcolor}

@@

\textit{odgi viz} supports a binned, linearized rendering in 1 dimension (1D). This visualization is computed in linear-time and offers a human-interpretable format suitable for understanding the topology and genome relationships in the pangenome graph (Fig. \ref{fig:odgi_viz}).
% Explain here odgi viz

Nevertheless, complex, nonlinear graph structures are difficult to display and interpret in a convenient low number of dimensions.
To overcome such a limitation, \textit{odgi viz} supports multiple visualization modes (Fig. \ref{fig:odgi_viz}\textbf{b-d}), making it easy to grasp the properties of the graph and its complexity.
% Explain here 3 odgi viz modes

@@

%\input{fig_odgi_viz}

>>>>

\usepackage[draft]{hyperref}
\usepackage{makecell}
\usepackage{comment}

% singlelinecheck=false puts subcaptions on the left
\usepackage[singlelinecheck=false]{subcaption}

\usepackage{algorithm2e}
\usepackage[usenames,dvipsnames]{xcolor}

@@

\textit{odgi viz} supports a binned, linearized rendering in 1 dimension (that is, all graph nodes lie on the same axis).
This visualization is computed in linear-time and offers a human-interpretable format suitable for understanding the topology and genome relationships in the pangenome graph (Fig. \ref{fig:odgi_viz}).
Graph nodes are arranged on a single axis, from left to right, with the colored bar indicating the paths and the nodes they cross.
White spaces indicate where paths do not traverse the nodes.
The meaning of the colors depends on how \textit{odgi viz} is executed.
By default, path colors has no particular meaning, but each path has its own color (Fig. \ref{fig:odgi_viz}\textbf{a}).
Path names are displayed on the left of the paths.
The black lines on the bottom indicate the edges connecting the nodes and, therefore, represent the graph topology.

Nevertheless, complex, nonlinear graph structures are difficult to display and interpret in a low number of dimensions.
To overcome such a limitation, \textit{odgi viz} supports multiple visualization modalities (Fig. \ref{fig:odgi_viz}\textbf{b-d}), making it easy to grasp the properties of the graph and its complexity.
Graph nodes order can affects downstream analyses on pangenome graphs. With \textit{odgi viz} we can color the paths by path position (Fig. \ref{fig:odgi_viz}\textbf{b}), with light grey indicating where paths begin and dark grey where they ends.
This visualization is suitable for checking graph nodes' linear order, as smooth color gradients indicate that the graph reflects well the linear paths' coordinate system.
Pangenome graphs represents both strand of DNA sequences.
\textit{odgi viz} supports also coloring the paths by orientation, with paths colored in red and black where their sequence is reversed or in forward, respectively, with respect to the sequences represented by the graph nodes (Fig. \ref{fig:odgi_viz}\textbf{b}).
Eukaryotic genomes experience gains and losses of genetic material, resulting in copy number variation (CNV) across the population.
With \textit{odgi viz}, we can use multiple color palettes to color the paths by path depth, highlighting the different copy number statuses in the genomes represented in the pangenome graph (Fig. \ref{fig:odgi_viz}\textbf{d}).

@@

\input{fig_odgi_viz}
